# Template

## Packages

```{r install_packages}
#| echo: true
#| eval: false

# install.packages('tidyverse')
```

```{r load_packages}
#| echo: true
#| eval: false

library(tidyverse)
```



## Constants

This is the place where you assign constants, e.g. the [Swiss VAT](https://www.estv.admin.ch/estv/en/home/value-added-tax/vat-rates-switzerland.html):

```{r constants}
#| echo: true
#| eval: false

VAT <- 0.077 # Swiss VAT (7.7%)
```

Remember that we use a different naming convention for constants, i.e. uppercase letters.



## Working Directories

RMarkdown files are a little special when it comes to working directories. But nevermind.

```{r wd}
#| echo: true
#| eval: false

getwd() # identify your current directory
setwd() # adjust it if needed
```


## Read Data

Ideally, you have a folder called `data` that contains two subfolders:

- `raw` from where you read the original data &
- `processed` where you store the data after you have transformed it

```{r read_data}
#| echo: true
#| eval: false

raw <- read.csv(file = 'data/raw/mtcars.csv',
                sep  = ',')
```

## Data Exploration

## Data Transformation

Do something with the data...

```{r data_manipulation}
#| echo: true
#| eval: false

processed <- raw
```



## Save Data

There are different formats you can use. CSV is the most compatible format.

```{r save_data}
#| echo: true
#| eval: false

write.csv(x = processed,
          file = 'data/processed/transformed_data.csv')

save(processed, 
     file = 'transformed_data.rda')
```



## Data Analyses

```{r data_analyses}

```

